---
title: "Diabetes130"
author: "Chris Tinley"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The dataset analyed for this project concerns hospital admissions for patients with diabetes at 130 different hospitals called, “Diabetes 130-US Hospitals for Years 1999-2008.” from the UC Irvine Machine Learning Repository. This is a rich dataset with 50 columns about the patient and their hospital stay.

I narrowed my focus to look at determinants of readmission and length of stay as these are important variables in terms of hospital economics. There are 717 unique diagnosis codes, so I narrowed the focus again down to four diagnoses (acute heart attack, hearth failure, pneumonia, coronary artery bypass graft) that are particularly salient because Centers for Medicare and Medicare (CMS) levies penalties on hospitals that have readmissions within 30 days for these diagnoses. I was also interested to see if discharge disposition was associated with readmission (ie., was the patient discharged home, home with services, to rehab). I looked at the breakdown for the sample itself and ran linear and multinomial analyses to figure out the most important factors contributing to length of stay and readmission status for these diagnoses.

## Part I – Exploring

```{r libaries, include=FALSE}
## Import Libaries ##
library(janitor)
library(lubridate)
library(tidyverse)
library(dplyr)
library(moments)
library(psych)
library(broom)
library(nnet)
library(reshape2)
library(rpart)
library(rpart.plot)
library(effects)
library(ggplot2)

```

```{r part 1, warning=FALSE}
## Load data from csv ##
v0_hospital_data <- read.csv("diabetic_data.csv")

### Part 1: Exploring ###
## Cleaning the data ##
# Race Column: Replace "?" with "unknown"
cleaned_hospital_data <- v0_hospital_data %>%
  mutate(column = recode(race, "?" = "unknown"))
# Age Column: Convert ranges to ordered factor for analysis
age_ranges = c("[0-10)","[10-20)","[20-30)","[30-40)","[40-50)","[50-60)", "[60-70)", "[70-80)", "[80-90)", "[90-100)")
age_ordered_factor <- factor(cleaned_hospital_data$age, 
  levels = c("[0-10)","[10-20)","[20-30)","[30-40)","[40-50)","[50-60)", "[60-70)", "[70-80)", "[80-90)", "[90-100)"), 
    ordered = TRUE)
# Gender Column: Replace "Unknown/Invalid" with "unknown"
cleaned_hospital_data <- cleaned_hospital_data %>%
  mutate(column = recode(gender, "Unknown/Invalid" = "unknown"))
# Weight Column: Replace "?" with "unknown"
cleaned_hospital_data <- cleaned_hospital_data %>%
  mutate(column = recode(weight, "?" = "unknown"))
# "admission_type_id" Column: Load reference table
admission_type_id_reference <- read.csv("admission_type_id.csv")
# "discharge_disposition_id" Column
  #Load reference table
    discharge_disposition_id_reference <- read.csv("discharge_disposition_id.csv")
  # Create and mapdischarge disposition groups
    dc_home <- c(1, 7, 16, 17, 19)
    dc_home_with_services <- c(6, 8, 13)
    dc_rehab <- c(3, 4, 15, 22, 24)
    dc_inpatient <- c(2, 5, 9,	10,	12,	20,	23,	27,	28,	29,	30)
    dc_other <- c(11,	14,	18,	21,	25,	26)
  # Create and map CMS readmission penalty diagnosis groups
    acute_MCI <- 410
    COPD <- c(491, 492, 496, 490)
    CHF <- c(428)
    Pneumonia <- c(480, 481, 482, 483, 484, 485, 486)
    CABG <- c(414)
    #OMIT THA_TKA due to lack of data <- c("V43")
# Create a new column for the discharge group
  cleaned_hospital_data <- cleaned_hospital_data %>%
    mutate(discharge_group = case_when(
      discharge_disposition_id %in% dc_home ~ "Home",
      discharge_disposition_id %in% dc_home_with_services ~ "Home with Services",
      discharge_disposition_id %in% dc_rehab ~ "Rehabilitation",
      discharge_disposition_id %in% dc_inpatient ~ "Inpatient Care",
      discharge_disposition_id %in% dc_other ~ "Other",
      TRUE ~ "Other"  # Catch any unexpected values
    ))
  # Create a new column for the diagnosis group
  cleaned_hospital_data <- cleaned_hospital_data %>%
    mutate(diagnosis_group = case_when(
      diag_1 %in% acute_MCI ~ "Acute Myocardial Infarction",
      diag_1 %in% CHF ~ "Congestive Heart Failure",
      diag_1 %in% Pneumonia ~ "Pneumonia",
      diag_1 %in% CABG ~ "Coronary Artery Bypass Graft",
      TRUE ~ "Other"  # Catch any unexpected values
    ))
  
  
    
## Descriptive statistics and graphs for interesting variables ##
# Age statistics
summary(age_ordered_factor)
age_ranges <- levels(age_ordered_factor)
age_counts <- summary(age_ordered_factor)
age_summary_df <- data.frame(
  age_ranges = age_ranges,
  count = as.numeric(age_counts))

ggplot(age_summary_df, aes(x = age_ranges, y = count)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Age Ranges", x = "Age Range", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Number of lab procedures statistics
describe(cleaned_hospital_data$num_lab_procedures)

ggplot(cleaned_hospital_data, aes(x = num_lab_procedures)) + 
  geom_histogram(bins = 20, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Number of Lab Procedures Per Patient", x = "Number of Lab Procedures", y = "Number of Patients")

# Number of medications statistics
describe(cleaned_hospital_data$num_medications)

ggplot(cleaned_hospital_data, aes(x = num_medications)) + 
  geom_histogram(bins = 20, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Number Medications Per Patient", x = "Number of Medications", y = "Number of Patients")

# Length of stay statistics
describe(cleaned_hospital_data$time_in_hospital)

ggplot(cleaned_hospital_data, aes(x = time_in_hospital)) + 
  geom_histogram(bins = 10, fill = "blue", alpha = 0.7) +
  scale_x_continuous(breaks=seq(min(cleaned_hospital_data$time_in_hospital), max(cleaned_hospital_data$time_in_hospital), by=1)) +
  labs(title = "Days in Hospital Per Patient", x = "Number of Days", y = "Number of Patients")

# Readmitted statistics
readmitted_summary <- cleaned_hospital_data %>%
  group_by(readmitted) %>%
  dplyr::summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)
readmitted_summary

ggplot(readmitted_summary, aes(x = "", y = percentage, fill = readmitted)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Readmission Status Distribution",
       fill = "Readmitted within 30 days",
       y = "Percentage") +
  theme_void() +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5))  # Add percentage labels

# Raw Data Associations - Age
age_readmitted_summary <- cleaned_hospital_data %>%
  group_by(age, readmitted) %>%
  dplyr::summarise(count = n(), .groups = 'drop') %>% 
  mutate(percentage = count / sum(count[readmitted == unique(readmitted)]) * 100)  # Calculate percentage for each age group

age_readmitted_summary_total <- cleaned_hospital_data %>%
  group_by(age, readmitted) %>%
  dplyr::summarise(count = n())

age_readmitted_summary

ggplot(age_readmitted_summary_total, aes(x = age, y = count, fill = readmitted)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = scales::comma(count)), 
            position = position_stack(vjust = 0.5), size = 2.5, color = "white", fontface = "bold") +  # Add percentage labels on bars
  labs(title = "Readmission Status by Age Group",
       x = "Age Group",
       y = "Proportion",
       fill = "Readmitted") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Raw Data Associations - Length of Stay
LOS_readmitted_summary <- cleaned_hospital_data %>%
  group_by(time_in_hospital, readmitted) %>%
  dplyr::summarise(count = n(), .groups = 'drop')
LOS_readmitted_summary_wide <- LOS_readmitted_summary %>%
  pivot_wider(names_from = readmitted, values_from = count)
LOS_readmitted_summary_wide

ggplot(LOS_readmitted_summary, aes(x = time_in_hospital, y = count, fill = readmitted)) +
  geom_bar(stat = "identity", position = "stack") +  geom_text(aes(label = scales::comma(count)), 
            position = position_stack(vjust = 0.5), size = 2.5, color = "white", fontface = "bold") +  # Add percentage labels on bars
  labs(title = "Readmission Status by Length of Stay",
       x = "Days in  Hospital",
       y = "Number of patients",
       fill = "Readmitted") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

  # by diagnosis
  avg_stay_by_diagnosis <- cleaned_hospital_data %>%
    group_by(diagnosis_group) %>%
    dplyr::summarise(avg_length_of_stay = mean(time_in_hospital, na.rm = TRUE), 
                     count = n(), 
                     .groups = 'drop') %>%
    arrange(desc(avg_length_of_stay))
  avg_stay_by_diagnosis
  
# Raw Data Associations - Number of procedures
summary_statistics_procedures <- cleaned_hospital_data %>%
  group_by(readmitted) %>%
  dplyr::summarise(
    min_procedures = min(num_procedures, na.rm = TRUE),
    max_procedures = max(num_procedures, na.rm = TRUE),
    mean_procedures = mean(num_procedures, na.rm = TRUE),
    median_procedures = median(num_procedures, na.rm = TRUE),
    sd_procedures = sd(num_procedures, na.rm = TRUE),
    count = n())
summary_statistics_procedures

ggplot(summary_statistics_procedures, aes(x = readmitted, y = mean_procedures, fill = readmitted)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Number of Procedures by Readmission Status",
       x = "Readmission Status",
       y = "Average Number of Procedures") +
  theme_minimal()

# Raw Data Associations - Number of medications
summary_statistics_meds <- cleaned_hospital_data %>%
  group_by(readmitted) %>%
  dplyr::summarise(
    min_meds = min(num_medications, na.rm = TRUE),
    max_meds = max(num_medications, na.rm = TRUE),
    mean_meds = mean(num_medications, na.rm = TRUE),
    median_meds = median(num_medications, na.rm = TRUE),
    sd_meds = sd(num_medications, na.rm = TRUE),
    count = n())
summary_statistics_meds

ggplot(summary_statistics_meds, aes(x = readmitted, y = mean_meds, fill = readmitted)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Number of Medications by Readmission Status",
       x = "Readmission Status",
       y = "Average Number of Medications") +
  theme_minimal()
```

## Part II: Expanding

```{r part ii, warning=FALSE}
## Discharge disposition and readmission
# Created 5 sub-categories for d/c disposition in Part 1, time to sum them up
  # Summarize the number of observations in each group
  discharge_summary <- cleaned_hospital_data %>%
    group_by(discharge_group) %>%
    dplyr::summarise(count = n())
  discharge_summary

  readmission_summary <- cleaned_hospital_data %>%
    group_by(discharge_group, readmitted) %>%
    dplyr::summarise(count = n(), .groups = 'drop')
    readmission_summary
  readmission_summary <- readmission_summary %>%
    group_by(discharge_group) %>%
    mutate(total_count = sum(count)) %>%
    mutate(percentage = (count / total_count) * 100) %>%
    ungroup()
  readmission_pivot <- readmission_summary %>%
    pivot_wider(names_from = readmitted, values_from = count, values_fill = 0)
  readmission_pivot
  
  # Stacked bar chart
  ggplot(readmission_summary, aes(x = discharge_group, y = count, fill = readmitted)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(aes(label = paste0(round(percentage, 1), "%")), 
    position = position_stack(vjust = 0.5), size = 3, color = "white") +
    labs(title = "Readmission Status by Discharge Disposition",
         x = "Discharge Disposition Group",
         y = "Number of Patients",
         fill = "Readmission Status") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

## Diagnosis and readmission
    discharge_summary2 <- cleaned_hospital_data %>%
    group_by(diagnosis_group) %>%
    dplyr::summarise(total_count = n())
  discharge_summary2
  
  readmission_summary2 <- cleaned_hospital_data %>%
    group_by(diagnosis_group, readmitted) %>%
    dplyr::summarise(count = n(), .groups = 'drop')
  readmission_summary2
  
  readmission_pivot2 <- readmission_summary2 %>%
    pivot_wider(names_from = readmitted, values_from = count, values_fill = 0)
  readmission_pivot2
  
  # Merge total counts with readmission summary
  readmission_pivot2 <- readmission_pivot2 %>%
    left_join(discharge_summary2, by = "diagnosis_group")
  readmission_pivot2
  
  readmission_long <- readmission_pivot2 %>%
    pivot_longer(cols = c(NO, `>30`, `<30`), names_to = "readmitted", values_to = "count") %>%
    mutate(percentage = (count / total_count) * 100)
  
  # Stacked bar chart
  ggplot(readmission_long, aes(x = diagnosis_group, y = count, fill = readmitted)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_x_discrete(limits = c("Acute Myocardial Infarction", "Congestive Heart Failure", "Coronary Artery Bypass Graft", "Pneumonia")) +
    geom_text(aes(label = paste0(round(percentage, 1), "%")), 
              position = position_stack(vjust = 0.5), size = 3, color = "white") +
    labs(title = "Readmission Status by Diagnosis Group",
         x = "Diagnosis Group",
         y = "Number of Patients",
         fill = "Readmission Status") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  
## Predicting length of stay
  # Convert categorical variables to factors
  cleaned_hospital_data$age_group <- as.factor(cleaned_hospital_data$age)
  cleaned_hospital_data$discharge_group <- as.factor(cleaned_hospital_data$discharge_group)
  
  # Linear model to predict time in hospital
  time_model <- lm(time_in_hospital ~ age + discharge_group + num_procedures + num_medications + diagnosis_group, data = cleaned_hospital_data)
  
  # Summary of the model
  summary(time_model)
  tidy_lm <- tidy(time_model, conf.int = TRUE)
  
  # Coefficient plot
  ggplot(tidy_lm, aes(x = estimate, y = term)) +
    geom_point() +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
    labs(title = "Coefficient Plot: Time in Hospital", x = "Estimate", y = "Predictor") +
    theme_minimal()
  

## Predicting readmission
  
  # Multinomial logistic regression model
  readmission_model <- multinom(readmitted ~ age + discharge_group + num_procedures + num_medications + diagnosis_group, 
                                data = cleaned_hospital_data)
  
  summary(readmission_model)
  
  diagnosis_effect <- effect("diagnosis_group", readmission_model)
  diagnosis_effect_df <- as.data.frame(diagnosis_effect)
  
  # Melt the data into long format for easier plotting
  melted_diagnosis_effect_df <- melt(diagnosis_effect_df, 
                                     id.vars = "diagnosis_group", 
                                     measure.vars = c("prob.X.30", "prob.X.30.1", "prob.NO"),
                                     variable.name = "readmission_status", 
                                     value.name = "probability")
  
  # Rename levels of readmission status for clarity
  melted_diagnosis_effect_df$readmission_status <- factor(melted_diagnosis_effect_df$readmission_status,
                                                          levels = c("prob.X.30", "prob.X.30.1", "prob.NO"),
                                                          labels = c("<30", ">30", "NO"))
  
  # Plot the predicted probabilities
  ggplot(melted_diagnosis_effect_df, aes(x = diagnosis_group, y = probability, fill = readmission_status)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_x_discrete(limits = c("Acute Myocardial Infarction", "Congestive Heart Failure", "Coronary Artery Bypass Graft", "Pneumonia")) +
    labs(title = "Predicted Probabilities by Diagnosis Group", 
         x = "Diagnosis Group", 
         y = "Predicted Probability") +
    scale_fill_brewer(palette = "Set2", name = "Readmission Status") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Part III: Communicating

I was interested in learning what factors contribute to readmission, and especially when the patient's diagnosis is targeted condition by the Centers for Medicare and Medicaid Services (CMS) and readmission could mean financial penalties for the hospital. I was also interested to learn about what factors increase length of stay, which is another variable that factors heavily into a hospital's finances.

My findings include:

-   The average number of lab procedures per patient during an admission is 43. The number of procedures is not strongly associated with length of stay or readmission status.

-   The average medications per patient is 16 and is also not strongly associated with length of stay or readmission

-   In this sample of patients with diabetes, 11.2% were readmitted within 30 days.

-   Of the key diagnoses, patients with congestive heart failure (CHF) seem to be the most concerning for readmission within 30 days. This diagnosis ranks the highest with 14.1% of patients in the sample readmitted within 30 days. With the statistical analysis, CHF diagnosis is also associated with the highest predicted probability of readmission within 30 days. The CHF diagnosis is suggested to increase length of stay somewhat (coefficient of 0.88). However, pneumonia was the diagnosis associated with longest length of stay both in the sample and the analysis.

-   From the discharge dispositions, patients recommended to discharge "Home" (without services or further inpatient care) had the highest percentage and gross amount of readmission within 30 days.